---
title: "Interactive visualisation"
author: "Quentin Geissmann"
date: "13 June 2017"
output: html_document
---


Aims
----- 

In this tutorial, we will learn:

* How to use `plotly` to display graphs 


The mariage `plotly` and `ggplot`
----------------------------------------

In `rethomics`, the graphs aere generated as `ggplot` objects. 
These can be exported as pdf which is great for publication-quality graph. 
However, sometimes, we want some degree of interactivenes with the figures, so we can zoom, exclude some data etc.
The packages `plotly` can take pretty much any ggplot object and make it interactive.

```{r, include=FALSE}
library(rethomics)
library(plotly)
source("rprint.R")
```



```{r, eval=FALSE}
install.packages("plotly")
library("plotly")
```



First, lets make an ethogram as a static figure [see here](ethogram_plot.html):

```{r, plot=TRUE}
data(sleep_sexual_dimorphism)
dt <- sleep_sexual_dimorphism
pl <- ethogramPlot(moving, dt, sex, error_bar="sem")
pl
```


This `pl` object is a static figure, so is is just displayed as an image.

Now by just using `ggplotly()` function, we can make the following:
```{r, plot=TRUE}
ggplotly(pl)
```

A few useful thing we can do here:

* By clicking on the legend, we can toggle out some groups
* By hovering on some point, we can get read `xy` values
* By selecting areas, we can zoom-in

For more info, have a look at the [plotly package](https://plot.ly/) itself.

