---
title: "Writing queries"
author: "Quentin Geissmann"
date: "1 March 2016"
output: html_document
---


Aims
----- 

In this tutorial, we will learn:

* The concept of query
* How some `rethomics` can use queries to load data

Queries are just like lab-book entries
----------------------------------------
This part is a bit abstract, but things will make more sense after reading the next few tutorials -- about specific kinds of data.

When doing an ethological experiment, you will most likely have multiple animals and several conditions/treatments/...
When you load large amounts of behavioural data, you generally want to retain information about biological conditions.
In other words, if you need a simple and flexible way to express the fact that you have applied a treatment to certain animals,
and you have to retrieve this variable in your [data table](data_structure.html).
This way, you can visualise and perform statistics in order to understand the effect of your treatments.

A query is a simple table where:

* each row is describes and animal
* each column is a condition

`rethomics` uses queries to locate and load data, so in most case queries will *compulsory columns* such as `region_id`, `path` to the result file...


An imaginary example of query
---------------------------------

```{r, include=FALSE}
library(rethomics)
```

As an example of simplistic query could, we could imagine an simple experiment with four animals that have a different sex and genotype:


```{r}
# our query is named q

q <- data.table(
   path="my_result_file.db",
   region_id = 1:4,
   sex=c("M","M","F","F"),
   genotype=c("A","B","A","B")
)
q
```

Here, you can see each row defines, unambiguously, an animal. Note that sometimes, you may have redundant variables.
In our example, the column `path` does not change because all data comes from the same result file.

According th the specific acquisition platform, you would use a different way to load the data, but the end goal is
always to use the query to map condition to time series.

Say our hypothetical acquisition platform simply generated 10 time point per animal and observed a single variable
(here random) named `x`, the resulting dt would look like:

```{r}
q[,
    .(t=1:10,
      x=round(runif(10),2)),
    by=eval(colnames(q))]
```


