---
title: "Position analysis"
author: "Quentin Geissmann"
date: "18 November 2015"
output: html_document
---

Aims
-----

This turorial demonstrates basic ways to visualise and analyse positional behavioural data.
We will learn how to:

* 
* `rethomics` use them to keep multiple animals in the same table.
* Work with data table keys.
* Use `rethomics` conventions.


Getting the data
-----------------------

```{r, include=FALSE}
library(rethomics)
```

We will use the `multiple_iterative_y_mazes` that is inside the package.

```{r}
data(multiple_iterative_y_mazes)

# we rename it to 'dt'
dt <- multiple_iterative_y_mazes
```

In this set of experiment, we have reccorded x,y,width,height, angle and other variable of individual fruit flies as they move through a maze.
The animals are males or females ("sex" column), and can bei either sleep deprived of not ("sleep_deprived" column).
In this data set, there is only one region of interest per experiment, so `region_id` is always 1.

First, we want to check how many animals we have per treatment/condition.
For that, we use some `data.table` trick:

```{r}
# this gives us only the first point for each animal
first_point_dt <- unique(dt)

# when we can get the number of animals:
nrow(first_point_dt)

# now we could like to count animals for of each sex and sleep deprivation treatment:
first_point_dt[, .N, by=.(sex, sleep_deprived)]
```


ggplot(dt, aes(x,y)) + geom_point() + geom_hex()

